---
title: "User Based Collaboorative Filtering"
output: html_notebook
---

## User based Collaborative Filtering

### Load ratings data, drop the "timestamp" column, and preview the data.

```{r}

# Define file path
ratings_path <- "/Users/vasista/Downloads/rating.csv"

# Load ratings data
ratings <- fread(ratings_path)

# Drop the timestamp column
ratings <- ratings[, !"timestamp", with=FALSE]

# Preview the ratings data
head(ratings)

```

### Load movies data and preview it

```{r}
# Define file path
movies_path <- "/Users/vasista/Downloads/movie.csv"

# Load movies data
movies <- fread(movies_path, quote="", fill=Inf)

# Preview the movie data
head(movies)

```

### Check the structure of the datasets.

```{r}
str(ratings)
str(movies)
```


```{r}
setnames(movies, gsub('\"', '', colnames(movies)))
```

### Merge datasets by 'movieId' and preview the merged data

```{r}
# Directly merge movies and ratings by 'movieId'
merged_data <- merge(ratings, movies, by = "movieId")

# View the merged data
head(merged_data)
```

### Calculate average ratings and rating counts for movies, and display the top-rated movies.

```{r}
# Calculate average ratings and rating counts for each movie
movie_avg_ratings <- merged_data %>%
  group_by(movieId, title) %>%
  summarise(
    avg_rating = mean(rating),
    rating_count = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(avg_rating))

# View the top 10 highest-rated movies
head(movie_avg_ratings, 10)

```

### Transform genres into binary columns, merge with average ratings, and preview the feature matrix.

```{r}
# Transform genres into binary columns
movie_features <- movies %>%
  separate_rows(genres, sep = "\\|") %>%
  mutate(value = 1) %>%
  pivot_wider(names_from = genres, values_from = value, values_fill = 0)

# Merge with average ratings
movie_features <- merge(movie_features, movie_avg_ratings, by = "movieId")

# View the feature matrix
head(movie_features)

```

### Create a binary genre feature matrix, prepare a numeric feature matrix, fit a Nearest Neighbors model, define a recommendation function, and test it.


```{r}
library(RANN)

# Transform movies to create a binary genre feature matrix
movie_features <- movies %>%
    separate_rows(genres, sep = "\\|") %>%
    mutate(value = 1) %>%
    pivot_wider(
        id_cols = c(movieId, title),  # Retain movieId and title
        names_from = genres,
        values_from = value,
        values_fill = 0
    )

# Verify column names
colnames(movie_features)

# Prepare the feature matrix (only numeric columns)
feature_matrix <- movie_features %>%
    select(where(is.numeric)) %>%
    as.matrix()

# Fit the Nearest Neighbors model
nn_model <- nn2(data = feature_matrix, k = 6)  # Find 5 nearest neighbors for each movie

# Function to get recommendations based on a movie ID
get_recommendations <- function(movie_id) {
    # Get the row index for the input movie
    row_index <- which(movie_features$movieId == movie_id)
    
    # Handle missing movie ID
    if (length(row_index) == 0) {
        stop("Movie ID not found in the dataset.")
    }
    
    # Find the nearest neighbors
    neighbors <- nn_model$nn.idx[row_index, -1]  # Exclude the input movie itself
    
    # Return recommended movie IDs and titles
    return(movie_features %>%
        filter(movieId %in% movie_features$movieId[neighbors]) %>%
        select(movieId, title))
}

# Test the recommendation function
recommendations <- get_recommendations(1)  # Replace '1' with a specific movie ID
print(recommendations)
```


```{r}
colnames(movie_features)

```

### Create a histogram to visualize the distribution of average movie ratings.

```{r}
# Visualize top-rated movies
library(ggplot2)

ggplot(movie_avg_ratings, aes(x = avg_rating)) +
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black") +
  labs(title = "Distribution of Average Movie Ratings", x = "Average Rating", y = "Count")

```

### Extract numeric columns from the movie_features dataset as a matrix.

```{r}
# Use only numeric columns from the movie_features dataset
X <- movie_features %>%
    select(where(is.numeric)) %>%
    as.matrix()

```


### Perform k-means clustering for different k values, calculate WCSS, and plot the elbow method.

```{r}
library(ggplot2)

# Define the dataset for clustering
X <- movie_features %>%
    select(where(is.numeric)) %>%
    as.matrix()

# Initialize an empty vector for within-cluster sum of squares (WCSS)
wcss <- c()
k_values <- 1:30  # Range of k values to test

# Perform k-means clustering for each k value
for (k in k_values) {
    kmeans_model <- kmeans(X, centers = k, nstart = 10)
    wcss <- c(wcss, kmeans_model$tot.withinss)
}

# Create a data frame for plotting
elbow_plot <- data.frame(k_values = k_values, wcss = wcss)

# Plot the elbow method
ggplot(elbow_plot, aes(x = k_values, y = wcss)) +
    geom_line() +
    geom_point() +
    labs(title = "The Elbow Method",
         x = "Number of Clusters (k)",
         y = "Within-Cluster Sum of Squares (WCSS)") +
    theme_minimal()

```


```{r}
```


```{r}
```


```{r}
```

